---
title: "HarvardX-PH125.9x: Choose Your Own Project Report"
subtitle: "Time series analysis using the example of weather data"
author: "Regina Castra"
date: "2019-06-07"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    latex_engine: xelatex
documentclass: article
classoption: a4paper
papersize: a4
fontsize: 11pt
mainfont: FreeSans   <!-- on linux machines FreeSans is an equivalent to Arial -->
urlcolor: blue
---

```{r LoadLibraries, include=FALSE, eval=TRUE}
### Load libraries and install them if not available yet
# reference: https://cran.r-project.org/web/views/TimeSeries.html
if(!require(colorspace)) install.packages("colorspace")
if(!require(tidyverse))  install.packages("tidyverse")
if(!require(readr))      install.packages("readr")
if(!require(caret))      install.packages("caret")
if(!require(knitr))      install.packages("knitr")
if(!require(kableExtra)) install.packages("kableExtra")
if(!require(munsell))    install.packages("munsell")
if(!require(scales))     install.packages("scales")
if(!require(ggplot2))    install.packages("ggplot2")
if(!require(reshape2))   install.packages("reshape2")
if(!require(latex2exp))  install.packages("latex2exp")
if(!require(xts))        install.packages("xts")
if(!require(corrplot))   install.packages("corrplot")
if(!require(ggrepel))    install.packages("ggrepel")
if(!require(grid))       install.packages("grid")
if(!require(forecast))   install.packages("forecast")
if(!require(lubridate))  install.packages("lubridate")
if(!require(quadprog))   install.packages("quadprog")
if(!require(tseries))    install.packages("tseries")
if(!require(zoo))        install.packages("zoo")
if(!require(xts))        install.packages("xts")
if(!require(ggfortify))  install.packages("ggfortify")
if(!require(forecast))   install.packages("forecast")
if(!require(nortest))    install.packages("nortest")


# define consistent ggplot-theme for all plots:
myggtheme <- theme_bw() +
  theme(plot.title   = element_text(size=9, face='bold'),
        axis.title.x = element_text(size=9, face='plain'),
        axis.title.y = element_text(size=9, face='plain'),
        axis.text.x  = element_text(size=8, face='plain'),
        axis.text.y  = element_text(size=8, face='plain'),
        legend.title = element_text(size=8, face='bold'),
        legend.text  = element_text(size=8, face='plain'))
colorsMM<- c(Mean="darkgreen", Median="red")


# define a standard function to create a ggplot of a histogram.
#   input: dataframe = a tibble of data, xData = data for histogram
myhistogram <- function(dataframe, xData, binwidth = 1, title = "", xLabel = "")
{
  median   <- median(xData)
  mean     <- mean(  xData)
  sd       <- sd(    xData)
  annotationText <- paste("Mean = "    , round(mean  , 3), "\n", 
                          "Median = "  , round(median, 3), "\n", 
                          "sigma = "   , round(sd    , 3), "\n",
                          "binwidth = ", binwidth, "\n",
                          "n = ", length(xData), sep="")
  ggplot(dataframe, aes(x = xData)) + 
    geom_histogram(binwidth = binwidth, color="darkblue", fill="lightblue", alpha=0.5) +
    geom_vline(aes(xintercept=median, color="Median"), size=1.5) +
    geom_vline(aes(xintercept=mean  , color="Mean"), size=1.5) +
    scale_colour_manual(name="Colors",values=colorsMM) +
    ggtitle(title) + 
    xlab(xLabel) +
    myggtheme +
    theme(legend.position = c(0.9, 0.8)) +  #x=0.85, y=0.845
    annotation_custom(grobTree(textGrob(annotationText, x=0.05, y=0.8, just="left", gp = gpar(fontsize=8) ))) + #y=0.85
    stat_function(fun = function(x) dnorm(x, mean = mean, sd = sd) * length(xData) * binwidth, color = "black", size = 1)
}

set.seed(123)

```


<!-- ------------------------------------------------------------------------------------------------------------------------------ -->
<!-- Instructions: (according to "Project Overview: Choose Your Own!")-->
<!-- For this project, you will be applying machine learning techniques that go beyond standard linear regression. -->
<!-- You will have the opportunity to use a publicly available dataset to solve the problem of your choice. -->
<!--   ... -->
<!-- The ability to clearly communicate the process and insights gained from an analysis is an important skill for data scientists. -->
<!--   ... -->
<!-- Although the exact format is up to you, the report should include the following at a minimum: -->
<!--   + an introduction/overview/executive summary section that describes the dataset and -->
<!--       summarizes the goal of the project and key steps that were performed; -->
<!--   + a methods/analysis section that explains the process and techniques used, such as data cleaning, data exploration and
<!--       visualization, any insights gained, and your modeling approach; -->
<!--   + a results section; and -->
<!--   + a conclusion section. -->
<!-- ------------------------------------------------------------------------------------------------------------------------------ -->


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




# Introductory Summary
The programming language **R** is considered particularly suitable for the analysis of time series,
as there are several packages and methods available to perform forecasts. 
For this project, the data of a weather station in Jena will be used to provide training and test dataset.

The goal is to train a machine learning algorithm using the daily mean temperatures of four years to predict the temperatures of the fifth year.
The quality of fitting the signal of the training dataset as well as the forecast's quality of the fifth year is quantified by a selection of accuracy measurements like RMSE, MAE, MASE and MAPE.

Starting with a base model which uses just the average as a constant forecast value, several different models are used.
A neural net method as well as models taking advantage of the seasonality of the data perform quite well in forecasting the temperatures.    



\pagebreak

# Analysis

## Data Source and Description

The used data file
[jena_climate_2009_2016.csv.zip](https://s3.amazonaws.com/keras-datasets/jena_climate_2009_2016.csv.zip)
is described on the
[website hosted at kaggle](https://www.kaggle.com/pankrzysiu/weather-archive-jena)
and can be downloaded from the
[Amazon Simple Storage Service (Amazon S3)](https://s3.amazonaws.com/keras-datasets/jena_climate_2009_2016.csv.zip)
without the need of any login credentials.
According to the
[website](https://www.kaggle.com/pankrzysiu/weather-archive-jena),
there were 14 different quantities (such as air temperature, atmospheric pressure, humidity, wind direction, and so on)
recorded every 10 minutes, over several years.
This dataset is limited to the years from 2009 to 2016. 




```{r DownloadData, include=FALSE, eval=TRUE}
# Download original file from https://s3.amazonaws.com  (no login credentials needed - very much appreciated for this project) 
#   https://s3.amazonaws.com/keras-datasets/jena_climate_2009_2016.csv.zip
dl <- tempfile()
download.file("https://s3.amazonaws.com/keras-datasets/jena_climate_2009_2016.csv.zip", dl)
```

The zip file provides only one csv-file:
```{r ZipFileContent, echo=FALSE}
unzip(dl, list=T) %>% knitr::kable() %>% 
  kable_styling(bootstrap_options = "striped", latex_options = "striped", full_width = FALSE, position = "center") %>% 
  row_spec(0, bold = TRUE)
```




## Data Structure
```{r CSVFileContentRaw, include=FALSE, eval=TRUE}
jena_climate_2009_2016_raw <- read_csv(unzip(dl, "jena_climate_2009_2016.csv"))
```

File *jena_climate_2009_2016.csv* has **`r ncol(jena_climate_2009_2016_raw)` attributes** stored in the columns 
and consist of **`r nrow(jena_climate_2009_2016_raw)` observations** in the rows.
The data seems to be in tidy format:

* Each variable in the dataset is placed in its own column
* Each observation is placed in its own row
* Each value is placed in its own cell

The 15 attributes have the following names, types and short descriptions (taken from the
[website hosted at kaggle](https://www.kaggle.com/pankrzysiu/weather-archive-jena)):

```{r DataAttributeTypes, echo=FALSE}
mydata.attributes <- sapply(jena_climate_2009_2016_raw, typeof)
mydata.attributes <- cbind(names(mydata.attributes),
                         mydata.attributes,
                         c("date & time",
                           "atmospheric pressure",
                           "temperature",
                           "potential temperature",
                           "dew point temperature",
                           "relative humidity",
                           "saturation water vapor pressure",
                           "actual water vapor pressure",
                           "water vapor pressure deficit",
                           "specific humidity",
                           "water vapor concentration",
                           "air density",
                           "wind velocity",
                           "maximum wind velocity",
                           "wind direction"))
mydata.attributes <- as_tibble(mydata.attributes, .name_repair = "minimal")
names(mydata.attributes) <- c("Attribute", "Type", "Description")
kable(mydata.attributes) %>% 
  kable_styling(bootstrap_options = "striped", latex_options = "striped", full_width = FALSE, position = "center") %>% 
  row_spec(0, bold = TRUE)
```

\pagebreak

To get a first impression of the data, the first 4 observations are shown in the following table.
Please note that the table has been transposed for better readability:

```{r dataFirst4, echo=FALSE}
mydata.first4 <- head(jena_climate_2009_2016_raw, n = 4)
mydata.first4 <- as_tibble(cbind(names(mydata.first4), t(mydata.first4)), .name_repair = "minimal")
names(mydata.first4) <- c("Attribute", "Observation 1", "Observation 2", "Observation 3", "Observation 4")
kable(mydata.first4) %>% 
  kable_styling(bootstrap_options = "striped", latex_options = "striped", full_width = FALSE, position = "center", font_size = 9.8) %>% 
  row_spec(0, bold = TRUE)
```




## Data Check and Transformation
The dataset does not contain any *missing values (NA)*. The number of *NA*-entries for each attribute in the dataset is: 
```{r dataCheckNA, echo=FALSE}
mydata.missingValues <- t(colSums(is.na(jena_climate_2009_2016_raw)))
row.names(mydata.missingValues) <- c("Number of NA")
kable(mydata.missingValues) %>%
  row_spec(0, angle = 90) %>% 
  kable_styling(bootstrap_options = "striped", latex_options = "striped", full_width = FALSE, position = "center", font_size = 9.8)
```

The statistical properties give a first impression of the data quality:
```{r dataStatistics, echo=FALSE}
kable(t(summary(jena_climate_2009_2016_raw))) %>% 
  kable_styling(bootstrap_options = "striped", latex_options = "striped", full_width = FALSE, position = "center", font_size = 9.5) %>% 
  row_spec(0, font_size = 0.1) %>% 
  add_header_above(c("Attribute" = 1, "Statistical Properties" = 6), bold = TRUE)
```

Comparing the "*Min.*" and "*Max.*" values with the quantiles "*1st Qu.*" and "*3rd Qu.*" lead to the conclusion that the values of the attributes "*wv (m/s)*" and "*max. wv (m/s)*" show outliers.

For the other attributes, 
[boxplots](https://www.r-bloggers.com/exploring-ggplot2-boxplots-defining-limits-and-adjusting-style/)
of their standardized values will give a graphical view of their distribution:  

```{r dataBoxplots, echo=FALSE, out.width="100%", fig.show="hold"}
mydata.standardized <- melt(as.data.frame(scale(select(jena_climate_2009_2016_raw, -c(`Date Time`, `max. wv (m/s)`, `wv (m/s)`)))),
                          id.vars = c(), variable.name = "Attribute", value.name = "StandardizedValue")
ggplot(mydata.standardized, aes(x = Attribute, y = StandardizedValue)) + geom_boxplot() +
  stat_boxplot(geom = "errorbar") + coord_flip() +
  ylab(TeX("$Standardized\\,Value\\;X' = \\frac{X - \\mu}{\\sigma}$")) +
  myggtheme
```
Attributes _"rho (g/m\*\*3)"_ and "_p (mbar)_" show separated *outliers*. The remaining attributes seem to have plausible values.


The first column "*Date Time*" is of type *character* and needs to be converted into a *POSIXct* "date time" column.
This is done by using the function *strptime()* with the format "%d.%m.%Y %H:%M:%S".

For analysis of time series it is mandatory to use data of constant lags.
According to the data description there should be an observation every 10 minutes.
A check confirms that the data has a periodicity of 10 minutes:
```{r checkPeriodicity, echo=FALSE}
# convert first column "Date Time" into a POSIXct date time format:
jena_climate_2009_2016 <- jena_climate_2009_2016_raw
jena_climate_2009_2016$`Date Time POSIXct` <- as.POSIXct(strptime(jena_climate_2009_2016$`Date Time`, format = "%d.%m.%Y %H:%M:%S", tz="UTC"))
# check periodicity of time series:
periodicity(jena_climate_2009_2016$`Date Time POSIXct`)
```

But there are also some observations missing (positive lag values > 10min) and others seem to be redundant (negative lag values).
For easier handling an unique *ID* is introduced for each observation.
The *ID* is equivalent to the line number in the *raw*-data.

```{r dataRawLags, echo=FALSE}
# add attribute ID and lag:
jena_climate_2009_2016 <- tibble::rowid_to_column(jena_climate_2009_2016, "ID")
jena_climate_2009_2016 <- jena_climate_2009_2016 %>% mutate(IDname = paste("ID", ID))
lags <- as.numeric(diff(jena_climate_2009_2016$`Date Time POSIXct`, lag = 1))
jena_climate_2009_2016$`lag (min)` <- c(0,lags)

# show the lags != 10min:
jena_climate_2009_2016 %>% select(-c(`Date Time POSIXct`)) %>% filter(`lag (min)` < 10 | 10 < `lag (min)`) %>%
  tail(-1) %>% select(ID, `Date Time`, `lag (min)`) %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped", latex_options = "striped", full_width = FALSE, position = "center")
# --> there are some observations missing (positive lag values) and others seem to be redundant (negative lag values)

# drop redundant values (78767 <= ID <= 78767+143) (redundant-range1)
jena_climate_2009_2016_clean <- subset(jena_climate_2009_2016, ID < 78767 | 78767+143 < ID)

# drop redundant values (274566 <= ID <= 274566+182) (redundant-range2)
jena_climate_2009_2016_clean <- subset(jena_climate_2009_2016_clean, ID < 274566 | 274566+182 < ID)

# calculate the lags again for the "clean" data:
lags <- as.numeric(diff(jena_climate_2009_2016_clean$`Date Time POSIXct`, lag = 1))
jena_climate_2009_2016_clean$`lag (min)` <- c(0,lags)
```

The first range of redundant values is shown in the following figure.
The resulting "cleaned" data is plotted as grey line.
The green line shows the observations up to *ID_78766* at time stamp *2010-07-02 00:00:00*.
The next observation *ID_78767* "jumps back" to time stamp *2010-07-01 00:10:00*.
This observation and subsequent ones are plotted with a blue line.

```{r showRedundantRange1, fig.align='center', echo=FALSE}
# for redundant-range1: visualize the drop-procedure and the resulting data w/o redundancy:
ggplot() +
  geom_line(data = subset(jena_climate_2009_2016_clean, 78767-300 < ID & ID < 78767+300),
            aes(x = `Date Time POSIXct`, y = `T (degC)`), color = "grey" , size=3.0) +
  geom_line(data = subset(jena_climate_2009_2016      , 78767-300 < ID & ID <= 78767    ),
            aes(x = `Date Time POSIXct`, y = `T (degC)`), color = "green", size=1.3) +
  geom_line(data = subset(jena_climate_2009_2016      , 78767     <= ID & ID < 78767+300),
            aes(x = `Date Time POSIXct`, y = `T (degC)`), color = "blue" , size=0.5) +
  xlab("Date Time") + ylab("T (degC)") +
  scale_x_datetime(labels = date_format("%Y-%m-%d\n%H:%M:%S", tz = "UTC")) +
  coord_cartesian(xlim = c(subset(jena_climate_2009_2016_clean,  ID == 78767-160) %>% select(`Date Time POSIXct`) %>% .[[1]],
                           subset(jena_climate_2009_2016_clean,  ID == 78767+160) %>% select(`Date Time POSIXct`) %>% .[[1]]),
                  ylim = c(subset(jena_climate_2009_2016,  78767-144 < ID & ID < 78767) %>% select(`T (degC)`) %>% min(),
                           subset(jena_climate_2009_2016,  78767-144 < ID & ID < 78767) %>% select(`T (degC)`) %>% max())) +
  geom_point(data = subset(jena_climate_2009_2016, ID %in% c(78767-144, 78767, 78766, 78767+143)),
             aes(x = `Date Time POSIXct`, y = `T (degC)`), shape = 21, size = 6, stroke = 1) +
  geom_label_repel(data = subset(jena_climate_2009_2016, ID %in% c(78767-144, 78767, 78766, 78767+143)),
                   aes(x = `Date Time POSIXct`, y = `T (degC)`, label = IDname),
                   box.padding = unit(1.2, 'lines'),
                   point.padding = unit(0.9, 'lines') ) +
  myggtheme 
```

The figure shows that the values are truely redundant, i.e. at each time stamp the green and the blue lines show the same value.
Same is true for the second range of redundant values:

```{r showRedundantRange2, fig.align='center', echo=FALSE}
# for redundant-range2: visualize the drop-procedure and the resulting data w/o redundancy:
ggplot() +
  geom_line(data = subset(jena_climate_2009_2016_clean, 274566-300 < ID & ID < 274566+300),
            aes(x = `Date Time POSIXct`, y = `T (degC)`), color = "grey" , size=3.0) +
  geom_line(data = subset(jena_climate_2009_2016      , 274566-300 < ID & ID <= 274566    ),
            aes(x = `Date Time POSIXct`, y = `T (degC)`), color = "green", size=1.3) +
  geom_line(data = subset(jena_climate_2009_2016      , 274566     <= ID & ID < 274566+300),
            aes(x = `Date Time POSIXct`, y = `T (degC)`), color = "blue" , size=0.5) +
  xlab("Date Time") + ylab("T (degC)") +
  scale_x_datetime(labels = date_format("%Y-%m-%d\n%H:%M:%S", tz = "UTC")) +
  coord_cartesian(xlim = c(subset(jena_climate_2009_2016_clean,  ID == 274566-190) %>% select(`Date Time POSIXct`) %>% .[[1]],
                           subset(jena_climate_2009_2016_clean,  ID == 274566+190) %>% select(`Date Time POSIXct`) %>% .[[1]]),
                  ylim = c(subset(jena_climate_2009_2016,  274566-183 < ID & ID < 274566) %>% select(`T (degC)`) %>% min(),
                           subset(jena_climate_2009_2016,  274566-183 < ID & ID < 274566) %>% select(`T (degC)`) %>% max())) +
  geom_point(data = subset(jena_climate_2009_2016, ID %in% c(274566-183, 274566, 274565, 274566+182)),
             aes(x = `Date Time POSIXct`, y = `T (degC)`), shape = 21, size = 6, stroke = 1) +
  geom_label_repel(data = subset(jena_climate_2009_2016, ID %in% c(274566-183, 274566, 274565, 274566+182)),
                   aes(x = `Date Time POSIXct`, y = `T (degC)`, label = IDname),
                   box.padding = unit(1.4, 'lines'),
                   point.padding = unit(0.9, 'lines') ) +
  myggtheme
```

After dropping the redundant values, the "cleaned" data looks at its joints as follows: 

```{r dataCleanedJoints, echo=FALSE}
# print data w/o redundant-range1:
subset(jena_climate_2009_2016_clean,  78767-3 < ID & ID <  78767+146) %>% select(ID, `Date Time`, `lag (min)`) %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped", latex_options = "striped", full_width = FALSE, position = "center")

# print data w/o redundant-range2:
subset(jena_climate_2009_2016_clean, 274566-3 < ID & ID < 274566+185) %>% select(ID, `Date Time`, `lag (min)`) %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped", latex_options = "striped", full_width = FALSE, position = "center")
```

The negative lags have been removed successfully from the data, but positive lags still remain:

```{r dataCleanedLags, echo=FALSE}
# show again the lags != 10min for the data without redundancy:
jena_climate_2009_2016_clean %>% select(-c(`Date Time POSIXct`)) %>% filter(`lag (min)` < 10 | 10 < `lag (min)`) %>%
  tail(-1) %>% select(ID, `Date Time`, `lag (min)`) %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped", latex_options = "striped", full_width = FALSE, position = "center")
```

These gaps usually need to be filled, as most modeling techniques require regular time series, i.e. they need a specific constant interval between observations.
The data will be aggregated using a time window of one hour, here by calculating a mean value for each hour.
Using this approach, the three gaps which are smaller than 60mins can be "closed".
Unfortunately two gaps over several hours still persist: 

```{r dataCleanedLags1h, echo=FALSE}
# aggregate the hourly mean:
jena_climate_2009_2016_aggHourly <- jena_climate_2009_2016_clean %>% 
  select(-c("ID", `Date Time`, "IDname", `lag (min)`)) %>% 
  mutate(`hour` = as.POSIXct(strptime(format(`Date Time POSIXct`, format = "%d.%m.%Y %H"), format = "%d.%m.%Y %H", tz="UTC"))) %>% 
  aggregate(list(`Date Time hour` = .$`hour`), mean) %>%
  mutate(`Date Time` = format(`Date Time hour`, format = "%d.%m.%Y %H:%M:%S")) %>% 
  select(-c(`Date Time POSIXct`, "hour")) %>% 
  mutate(`lag (h)` = c(0,as.numeric(diff(`Date Time hour`, lag = 1)))) %>% 
  tibble::rowid_to_column("IDh")
# show the lags != 1h:
jena_climate_2009_2016_aggHourly %>% select(-c(`Date Time hour`)) %>% tail(-1) %>% filter(`lag (h)` < 1 | 1 < `lag (h)`) %>%
  select(IDh, `Date Time`, `lag (h)`) %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped", latex_options = "striped", full_width = FALSE, position = "center")
```

The small gaps of 20 and 30 minutes disappeared.
The following figure gives an impression, how this was accomplished.
The raw-data (black line) is used to create a mean value for each hour (blue line).
There are two observations missing between *ID_40378* and *ID_40379* (grey band).
By averaging over 1h intervalls the gap is automatically "closed".
Remark: Attribute *IDh* is introduced which provides an unique integer for each hourly observation.

```{r aggregateCloseGap, fig.align='center', echo=FALSE}
# plot raw- and aggregated-data at the place were a lag=30min was detected:
ggplot() +
  geom_line(data = subset(jena_climate_2009_2016_clean, 40379-1 <= ID & ID <= 40379),
            aes(x = `Date Time POSIXct`, y = `T (degC)`), color = "grey" , size=5) +
  geom_line(data = subset(jena_climate_2009_2016_clean, round(40379-0.2*24*6) < ID & ID <= 40379-1),
            aes(x = `Date Time POSIXct`, y = `T (degC)`), color = "black" , size=1) +
  geom_line(data = subset(jena_climate_2009_2016_clean, 40379 <= ID & ID < round(40379+0.2*24*6)),
            aes(x = `Date Time POSIXct`, y = `T (degC)`), color = "black" , size=1) +
  geom_point(data = subset(jena_climate_2009_2016_clean, round(40379-0.2*24*6) < ID & ID < round(40379+0.2*24*6)),
            aes(x = `Date Time POSIXct`, y = `T (degC)`), shape = 21, color = "black" , fill = "white" , size=2.5) +
  xlab("Date Time") + ylab("T (degC)") +
  scale_x_datetime(labels = date_format("%Y-%m-%d\n%H:%M:%S", tz = "UTC")) +
  coord_cartesian(xlim = c(subset(jena_climate_2009_2016_clean,  ID == round(40379-0.1*24*6)) %>% select(`Date Time POSIXct`) %>% .[[1]],
                           subset(jena_climate_2009_2016_clean,  ID == round(40379+0.1*24*6)) %>% select(`Date Time POSIXct`) %>% .[[1]]),
                  ylim = c(subset(jena_climate_2009_2016, round(40379-0.1*24*6) < ID & ID < round(40379+0.1*24*6)) %>% select(`T (degC)`) %>% min(),
                           subset(jena_climate_2009_2016, round(40379-0.1*24*6) < ID & ID < round(40379+0.1*24*6)) %>% select(`T (degC)`) %>% max())) +
  geom_step(data = jena_climate_2009_2016_aggHourly,
            aes(x = `Date Time hour`, y = `T (degC)`), color = "blue", alpha = 0.6, size=1) +
  geom_point(data = subset(jena_climate_2009_2016_clean, ID %in% c(40379-1, 40379)),
             aes(x = `Date Time POSIXct`, y = `T (degC)`), shape = 21, size = 6, stroke = 1) +
  geom_label_repel(data = subset(jena_climate_2009_2016, ID %in% c(40379-1, 40379)),
                   aes(x = `Date Time POSIXct`, y = `T (degC)`, label = IDname),
                   box.padding = unit(1.1, 'lines'),
                   point.padding = unit(0.9, 'lines') ) +
  myggtheme
```

Following plot illustrates the aggregation approach and shows how well the 1h-average follows the original data.
It can be thought of as a noise-reduction, i.e. higher frequencies are damped.

```{r aggregateFollowsRaw, fig.align='center', echo=FALSE, fig.height=4.5, out.width="100%", fig.show="hold"}
# plot raw- and aggregated-data for over ca. 3 days:
ggplot() +
  geom_line(data = subset(jena_climate_2009_2016_clean, round(40379-3*24*6) < ID & ID < round(40379+3*24*6)),
            aes(x = `Date Time POSIXct`, y = `T (degC)`), color = "black" , size=0.5) +
  xlab("Date Time") + ylab("T (degC)") +
  scale_x_datetime(labels = date_format("%Y-%m-%d\n%H:%M:%S", tz = "UTC")) +
  coord_cartesian(xlim = c(subset(jena_climate_2009_2016_clean,  ID == round(40379-1.5*24*6)) %>% select(`Date Time POSIXct`) %>% .[[1]],
                           subset(jena_climate_2009_2016_clean,  ID == round(40379+1.6*24*6)) %>% select(`Date Time POSIXct`) %>% .[[1]]),
                  ylim = c(subset(jena_climate_2009_2016, round(40379-1.5*24*6) < ID & ID < round(40379+1.6*24*6)) %>% select(`T (degC)`) %>% min(),
                           subset(jena_climate_2009_2016, round(40379-1.5*24*6) < ID & ID < round(40379+1.6*24*6)) %>% select(`T (degC)`) %>% max())) +
  geom_step(data = jena_climate_2009_2016_aggHourly,
            aes(x = `Date Time hour`, y = `T (degC)`), color = "blue", alpha = 0.6, size=1) +
  myggtheme 
```

The resulting data used in the following chapters starts at "01.01.2009_00:00:00" and will be cut off at "24.09.2014_00:00:00".

```{r dataCutOff_2009_2014, echo=FALSE}
# cut the data off at 24.09.2014 00:00:00 (inclusive)
jena_climate_2009_2014h <- jena_climate_2009_2016_aggHourly %>%
  subset(`Date Time hour` < as.POSIXct("2014-09-24 00:00:01", tz = "UTC")) %>% 
  select(-c(`lag (h)`))

# clean up:
rm(jena_climate_2009_2016, jena_climate_2009_2016_clean, jena_climate_2009_2016_raw,
   dl, lags, mydata.first4, mydata.missingValues, mydata.standardized,
   jena_climate_2009_2016_aggHourly, mydata.attributes)
```




## Data Analysis
The hourly data is clean for the time period "01.01.2009_00:00:00" to "24.09.2014_00:00:00".
In the following the standard pre-processing and analysis for regular time series will be performed
mainly on the attribute "*T (degC)*", i.e. the focus will be on univariate regular time series modelling.

During the first investigations it became obvious that working on hourly data requires a lot of computing resources.
Therefore an aggregation is performed again to obtain daily data, i.e. for the given time period using the hourly data,
the mean temperature "*T (degC)*" for each day is calculated.

```{r aggregateDataDailyMonthly, include=FALSE}
# reduce the clean data to the time range (01.01.2009_00:00:00-31.01.2013_23:00:00) to obtain a dataset with frequency of 365 days (whole year)
jena_climate_2009_2013h <- jena_climate_2009_2014h %>% subset(`Date Time hour` < as.POSIXct("2014-01-01 00:00:00", tz = "UTC"))
head(jena_climate_2009_2013h)
tail(jena_climate_2009_2013h)  

# it turned out during the investigations that the hourly data requires quite a lot of computing resources...
# therefore another aggregation for daily mean data and monthly mean data (based on hourly data) is performed:
# aggregate the daily mean:
jena_climate_2009_2013d <- jena_climate_2009_2013h %>% 
  select(-c("IDh", `Date Time`)) %>% 
  mutate(`day` = as.POSIXct(strptime(format(`Date Time hour`, format = "%d.%m.%Y"), format = "%d.%m.%Y", tz="UTC"))) %>% 
  aggregate(list(`Date Time day` = .$`day`), mean) %>%
  mutate(`Date Time` = format(`Date Time day`, format = "%d.%m.%Y %H:%M:%S")) %>% 
  select(-c(`Date Time hour`, "day")) %>% 
  tibble::rowid_to_column("IDd")
jena_climate_2009_2013d
head(jena_climate_2009_2013d)
tail(jena_climate_2009_2013d)
summary(jena_climate_2009_2013d)

# aggregate the monthly mean:
jena_climate_2009_2013m <- jena_climate_2009_2013h %>% 
  select(-c("IDh", `Date Time`)) %>% 
  mutate(`month` = as.POSIXct(strptime(paste("01.",format(`Date Time hour`, format = "%m.%Y")), format = "%d.%m.%Y", tz="UTC"))) %>% 
  aggregate(list(`Date Time month` = .$`month`), mean) %>%
  mutate(`Date Time` = format(`Date Time month`, format = "%d.%m.%Y %H:%M:%S")) %>% 
  select(-c(`Date Time hour`, "month")) %>% 
  tibble::rowid_to_column("IDm")
jena_climate_2009_2013m
head(jena_climate_2009_2013m)
tail(jena_climate_2009_2013m)
summary(jena_climate_2009_2013m)
```

Furthermore the time period of the investigated data is reduced to five years (01.01.2009_00:00-31.01.2013_23:00)
to obtain a dataset with frequency of 365 days (whole year).
For the hourly, daily and monthly aggregated data the histograms are presented: 

```{r histogramHDM, fig.align='center', echo=FALSE, fig.height=3.4, out.width="100%", fig.show="hold"}
# show histogram of hourly, daily, monthly data
myhistogram(jena_climate_2009_2013h, jena_climate_2009_2013h$`T (degC)`, binwidth = 1, xLabel = "T (degC)") +
  ggtitle("Distribution of the hourly \"T (degC)\" in the period 01.01.2009 00:00:00 - 31.01.2013 23:00:00")+
  coord_cartesian(xlim = c(-20,30))
myhistogram(jena_climate_2009_2013d, jena_climate_2009_2013d$`T (degC)`, binwidth = 2, xLabel = "T (degC)") +
  ggtitle("Distribution of the daily \"T (degC)\" in the period 01.01.2009 - 31.01.2013") +
  coord_cartesian(xlim = c(-20,30))
myhistogram(jena_climate_2009_2013m, jena_climate_2009_2013m$`T (degC)`, binwidth = 3, xLabel = "T (degC)") +
  ggtitle("Distribution of the monthly \"T (degC)\" in the period Jan. 2009 - Dec. 2013") +
  coord_cartesian(xlim = c(-20,30))
# --> calculating the mean has a huge influence on distribution, i.e. big difference to normal distribution 
```

It turns out, that calculating the mean has a huge influence on the distribution:
By averaging over a longer period of time the distribution deviates more and more from normal distribution.
Normality test methods like the 
[Shapiro-Wilk test](https://en.wikipedia.org/wiki/Shapiro%E2%80%93Wilk_test)
and the
[Anderson-Darling test](https://en.wikipedia.org/wiki/Anderson%E2%80%93Darling_test)
applied to the hourly, daily and monthly aggregated data also support this theses by returning higher *p-values* for increasing deviation:  
<!-- -->
```{r tableNormalityTest, fig.align='center', echo=FALSE, out.width="100%", fig.show="hold"}
normalityTestTb <- tibble('hourly data'  = "N/A", 
                          'daily data'   = format(shapiro.test(jena_climate_2009_2013d$`T (degC)`)$p.value, digits = 3),
                          'monthly data' = format(shapiro.test(jena_climate_2009_2013m$`T (degC)`)$p.value, digits = 3))
normalityTestTb <- add_row(normalityTestTb,  'hourly data'  = format(ad.test(jena_climate_2009_2013h$`T (degC)`)$p.value, digits = 3),
                                             'daily data'   = format(ad.test(jena_climate_2009_2013d$`T (degC)`)$p.value, digits = 3),
                                             'monthly data' = format(ad.test(jena_climate_2009_2013m$`T (degC)`)$p.value, digits = 3))
normalityTestTb <- add_column(normalityTestTb, 'Test on data' = c("Shapiro-Wilk normality test, p-value", "Anderson-Darling normality test, p-value"), .before = 1)
kable(normalityTestTb) %>% kable_styling(bootstrap_options = "striped", latex_options = "striped", full_width = FALSE, position = "center")
rm(normalityTestTb)
```
<!-- -->
Aggregating reduces the maximum values (spikes) like a filter, which is well shown by line plots of "T (degC)":

```{r visualizeByLinePlotsHDM, fig.align='center', echo=FALSE, fig.height=4.2, out.width="100%", fig.show="hold"}
# create time-series objects (ts) for the given data (hourly, daily, monthly):
tsJenaTh <- ts(data = jena_climate_2009_2013h$`T (degC)`, start = c(2009,0), frequency = 24*365)
tsJenaTd <- ts(data = jena_climate_2009_2013d$`T (degC)`, start = c(2009,0), frequency = 365)
tsJenaTm <- ts(data = jena_climate_2009_2013m$`T (degC)`, start = c(2009,1), frequency = 12)
tsJenaTh <- window(tsJenaTh, start=c(2009,0.001), end=c(2013,24*365-4)) # because of some reason there are 2008 and 2014-entries!!!
tsJenaTd <- window(tsJenaTd, start=c(2009,0.001), end=c(2013,365)) # because of some reason there are 2008 and 2014-entries!!!

# plot line charts of the 3 datasets:
ggplot()+
  geom_line(data = data.frame(date=time(tsJenaTh)        , TempC=as.matrix(tsJenaTh)), aes(x = date, y = TempC, color = "hourly")) +
  geom_line(data = data.frame(date=time(tsJenaTd)+0.5/365, TempC=as.matrix(tsJenaTd)), aes(x = date, y = TempC, color = "daily")) + 
  geom_line(data = data.frame(date=time(tsJenaTm)+0.5/12 , TempC=as.matrix(tsJenaTm)), aes(x = date, y = TempC, color = "monthly"), size=1) +
  xlab("Date") +
  ylab("T (degC)") +
  ggtitle("Values of \"T (degC)\" in the time period of 01.01.2009 00:00:00 to 01.01.2014 00:00:00 (excl.)") +
  myggtheme +  
  theme(legend.direction = "horizontal", legend.position = c(0.66, 0.06)) +
  scale_color_manual(name = "Data aggregation (mean value)",
                     breaks = c("hourly", "daily", "monthly"),
                     values = c("hourly" = "darkgrey", "daily" = "blue", "monthly" = "black") ) +
  scale_x_continuous()
```

The plot shows data of high seasonality with a yearly frequency, i.e. every 12 months the signal seems to repeat itself.
Seasonal plots (see [function seasonplot()](https://rdrr.io/cran/forecast/man/seasonplot.html))
are able to visualize this behaviour by plotting the data against the seasons (January to December) for separate years, see also [Hyndman (2014, chapter 2)](https://robjhyndman.com/uwafiles/fpp-notes.pdf):

```{r seasonalPlotHDM, fig.align='center', echo=FALSE, fig.height=3.1, out.width="100%", fig.show="hold"}
suppressMessages(ggseasonplot(tsJenaTh) + myggtheme +
  ggtitle("Seasonal plot of hourly \"T (degC)\" for the years 2009 to 2013") +
    scale_y_continuous(name="T (degC)", limits=c(-25, 35), breaks=c(-25,-20,-15,-10,-5,0,5,10,15,20,25,30,35)) +
    scale_x_continuous(name="Month",
                       breaks = ((1:12)-0.5)/12,
                       labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")))

suppressMessages(ggseasonplot(tsJenaTd) + myggtheme + 
  ggtitle("Seasonal plot of daily \"T (degC)\" for the years 2009 to 2013") +
    scale_y_continuous(name="T (degC)", limits=c(-25, 35), breaks=c(-25,-20,-15,-10,-5,0,5,10,15,20,25,30,35)) +
    scale_x_continuous(name="Month",
                       breaks = ((1:12)-0.5)/12,
                       labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")))

suppressMessages(ggseasonplot(tsJenaTm)+ myggtheme +
  ggtitle("Seasonal plot of monthly \"T (degC)\" for the years 2009 to 2013") +
    scale_y_continuous(name="T (degC)", limits=c(-25, 35), breaks=c(-25,-20,-15,-10,-5,0,5,10,15,20,25,30,35)) +
    scale_x_continuous(name="Month",
                       breaks = ((1:12)-0.5)/13,
                       labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")))
```

To get a better impression, how much the value of "T (degC)" varies from year to year for each hour/day/month,
so called "month plots" 
(see [function monthplot()](https://www.rdocumentation.org/packages/stats/versions/3.6.0/topics/monthplot)) are created.
For each hour/day/month of the five given years, the value is plotted side by side together with its average as blue line.  

```{r monthPlotH, fig.align='center', echo=FALSE, fig.height=2.9, out.width="100%", fig.show="hold"}
suppressMessages(ggmonthplot(tsJenaTh) + myggtheme +
  ggtitle("\"Month\" (=hour) plot of hourly \"T (degC)\" for the years 2009 to 2013") +
    scale_y_continuous(name="T (degC)", limits=c(-25, 35), breaks=c(-25,-20,-15,-10,-5,0,5,10,15,20,25,30,35)) +
    scale_x_continuous(name="Month",
                     breaks = (0:11)*365*2 + 365,
                     labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")))
```
```{r monthPlotD, fig.align='center', echo=FALSE, fig.height=2.9, out.width="100%", fig.show="hold"}
suppressMessages(ggmonthplot(tsJenaTd) + myggtheme +
  ggtitle("\"Month\" (=day) plot of daily \"T (degC)\" for the years 2009 to 2013") +
    scale_y_continuous(name="T (degC)", limits=c(-25, 35), breaks=c(-25,-20,-15,-10,-5,0,5,10,15,20,25,30,35)) +
    scale_x_continuous(name="Month",
                       breaks = (0:11)*365/12 + 365*0.5/11,
                       labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")))
```
```{r monthPlotM, fig.align='center', echo=FALSE, fig.height=2.9, out.width="100%", fig.show="hold"}
suppressMessages(ggmonthplot(tsJenaTm) + myggtheme +
  ggtitle("Month plot of monthly \"T (degC)\" for the years 2009 to 2013") +
    scale_y_continuous(name="T (degC)", limits=c(-25, 35), breaks=c(-25,-20,-15,-10,-5,0,5,10,15,20,25,30,35)) +
    scale_x_continuous(name="Month",
                       breaks = (1:12)+0.5,
                       labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")))
```

The plots confirm data of high seasonality.
This characteristic can be used for forecasting.
Most forecasting methods depend on data without a trend present. 
By decomposing the data regarding season and trend, the remainder does not show any obvious pattern:

```{r decomposeH, fig.align='center', echo=FALSE, fig.height=4.4, out.width="100%", fig.show="hold", message=FALSE, warning=FALSE}
suppressWarnings(autoplot(decompose(tsJenaTh, type = "additive")) + myggtheme + 
                   ggtitle("Decompose plots of hourly \"T (degC)\" for the years 2009 to 2013") ) +
                   xlab("Date")
```

```{r decomposeD, fig.align='center', echo=FALSE, fig.height=4.4, out.width="100%", fig.show="hold", message=FALSE, warning=FALSE}
suppressWarnings(autoplot(decompose(tsJenaTd, type = "additive")) + myggtheme + 
                   ggtitle("Decompose plots of daily \"T (degC)\" for the years 2009 to 2013") ) +
                   xlab("Date")
```

```{r decomposeM, fig.align='center', echo=FALSE, fig.height=4.4, out.width="100%", fig.show="hold", message=FALSE, warning=FALSE}
suppressWarnings(autoplot(decompose(tsJenaTm, type = "additive")) + myggtheme  + 
                   ggtitle("Decompose plots of monthly \"T (degC)\" for the years 2009 to 2013") ) +
                   xlab("Date")
```

The [Augmented Dickey–Fuller test](https://en.wikipedia.org/wiki/Augmented_Dickey%E2%80%93Fuller_test)
delivers *p-values* smaller than 0.015, which is also an indication for stationary data, i.e. data without a trend. 
According to 
[Tavish Srivastava's Time Series Tutorial](https://www.analyticsvidhya.com/blog/2015/12/complete-tutorial-time-series-modeling/)
stationary data must fulfill the following:

1. The mean of the series should be constant over time.

2. The variance of the series should be constant over time.

3. The covariance of the series should be constant over time.


**Summary of data analysis**

By looking at the line plots of the data and its decomposition it can be summarized:  
The data is of high seasonality, is stationary and shows no distinct trend.
The remainder has no obvious pattern.




\pagebreak

# Modeling
In this chapter training and test dataset will be defined.
After introducing the methods to quantify the accuracy, several models for forecasting will be presented.
 
## Train and Test Dataset
In order to evaluate the quality of the forecast, the daily data of the 5-year period is split into two datasets:
The time frame 01.01.2009-31.12.2012 will be used as training dataset.
The values of the fifth year (2013) are sepcified as test dataset (holdout) and will be compared to the forecasted values. 
The following line plot shows the two datasets:

```{r linePlotTrainingTest, fig.align='center', echo=FALSE, fig.height=3.5, out.width="100%", fig.show="hold"}
# Devide daily data into training and testing dataset: 80% for training (01.01.2009-31.12.2012), holdout of 20% for testing (01.01.2013-31.01.2013)
# Whole years are selected
tsJenaTd.train <- window(tsJenaTd, start=c(2009,1), end=c(2012,365))
tsJenaTd.test  <- window(tsJenaTd, start=c(2013,1), end=c(2013,365))
tsJenaTm.train <- window(tsJenaTm, start=c(2009,1), end=c(2012,12))
tsJenaTm.test  <- window(tsJenaTm, start=c(2013,1), end=c(2013,12))

# Plot train- and test-dataset:
ggplot() +
  forecast::autolayer(tsJenaTd.train, series = 'train data') +
  forecast::autolayer(tsJenaTd.test , series = 'test data') +
  xlab('Year') + ylab('daily mean of \"T (degC)\"') + 
  guides(colour = guide_legend(title = 'Dataset', reverse=T)) +
  myggtheme +
  coord_cartesian(ylim = c(-25,40)) +
  theme(legend.position = c(0.88, 0.14)) +
  scale_color_manual(values=c("green", "black")) 
```

A first comparison of the two datasets shows very different distributions to each other.
Both datasets deviate from normality distribution:

```{r distributionTraining, fig.align='center', echo=FALSE, fig.height=3.4, out.width="100%", fig.show="hold"}
# Plot histogram of train dataset "T (degC)" (daily data):
tsJenaTd.train.df <- data.frame(TdegC = as.matrix(tsJenaTd.train), date = time(tsJenaTd.train))
myhistogram(tsJenaTd.train.df, tsJenaTd.train.df$TdegC, binwidth = 2, xLabel = "T (degC)") +
  ggtitle("Training dataset: Distribution of the daily \"T (degC)\" for the years 2009-2012") +
  coord_cartesian(xlim = c(-20,30))
# --> has some deviations to a normal distribution 
```

```{r distributionTest, fig.align='center', echo=FALSE, fig.height=3.4, out.width="100%", fig.show="hold"}
# Plot histogram of test dataset "T (degC)" (daily data):
tsJenaTd.test.df <- data.frame(TdegC = as.matrix(tsJenaTd.test), date = time(tsJenaTd.test))
myhistogram(tsJenaTd.test.df, tsJenaTd.test.df$TdegC, binwidth = 2, xLabel = "T (degC)") +
  ggtitle("Test dataset: Distribution of the daily \"T (degC)\" for the year 2013") +
  coord_cartesian(xlim = c(-20,30))
# --> does not look like a normal distribution and also quite different to the training dataset

# introduce a table for collecting all accuracies of the models:
accuracyTb.train <- tibble() # for collecting all accuracies of the models at the training dataset
accuracyTb.test  <- tibble() # for collecting all accuracies of the models at the test dataset
```


## Quantification of Accuracy
The R-function *accuracy()* is used to quantify the quality of the different models regarding fitting the *training* dataset and forecasting the *test* datatset.
Following selection of methods is used for comparison:

* **MAE: Mean Absolute Error** - mean of all differences between actual and forecasted absolute values, see
[Wikipedia-MAE](https://en.wikipedia.org/wiki/Mean_absolute_error)

* **RMSE: Root Mean Squared Error** - the sample standard deviation of differences between actual and forecasted values, see
[Wikipedia-RMSE](https://en.wikipedia.org/wiki/Root-mean-square_deviation)

* **MASE: Mean Absolute Scaled Error** - measures the mean absolute error of the forecast values, divided by the mean absolute error of the *naive* forecast.
[Wikipedia-MASE](https://en.wikipedia.org/wiki/Mean_absolute_scaled_error)

* **MAPE: Mean Absolute Percentage Error** - measures the difference of forecast errors and divides it by the actual value, see
[Wikipedia-MAPE](https://en.wikipedia.org/wiki/Mean_absolute_percentage_error)




<!-- --------------------------------------------------------------------------------------------- -->
\pagebreak 

## Model 1: Naive Method
```{r m1Forecast, echo=FALSE}
tsJenaTd.fc_naive  <-  naive(tsJenaTd.train, h=365)
```
The naive model projects the last observation into the future.
So all forecasts are constant and equal to the last observed value.
Here in this case the predicted constant value for the year 2013 is `r tsJenaTd.fc_naive$mean[1]`.


The prediction provided by this model for the year 2013 is shown as blue line: 

```{r m1PlotForecast, fig.align='center', echo=FALSE, fig.height=2.8, out.width="100%", fig.show="hold"}
autoplot(tsJenaTd.fc_naive) + myggtheme +
  xlab("Date") +
  ylab("T (degC)") +
  ggtitle("Model 1: Naive method, forecast of \"T (degC)\"") +
  coord_cartesian(ylim = c(-25,40))
```

With this model the following accuracies on training and test dataset can be achieved: 

```{r m1TableAccuracy, fig.align='center', echo=FALSE, out.width="100%", fig.show="hold"}
# show accuracy of forecast model:
accuracy(tsJenaTd.fc_naive , tsJenaTd.test) %>%
  kable() %>% kable_styling(bootstrap_options = "striped", latex_options = "striped", full_width = FALSE, position = "center")

# gather accuracies of forecast models:
accuracyTb.train <- bind_rows(accuracyTb.train, add_column(as_tibble(t(accuracy(tsJenaTd.fc_naive , tsJenaTd.test)[1,])), Model="M1 naive"         , .before = 1))
accuracyTb.test  <- bind_rows(accuracyTb.test , add_column(as_tibble(t(accuracy(tsJenaTd.fc_naive , tsJenaTd.test)[2,])), Model="M1 naive"         , .before = 1))
```

The residual distribution of this model is shown by a histogram:

```{r m1PlotResiduals, fig.align='center', echo=FALSE, fig.height=2.8, out.width="100%", fig.show="hold"}
# Plot residuals
res.df <- na.omit(data.frame(res = as.matrix(tsJenaTd.fc_naive$residuals )))
myhistogram(res.df, res.df$res, 1, "Residuals of model 1: naive method"  , "Residuals") + coord_cartesian(xlim = c(-25,20))
```

The residuals tell a lot about the quality of the model.
They are what is left after the fitted values (here it is only one constant value) got subtracted from the original dataset.
Only randomness should stay in the residuals. Everything else should be included in the model.
Therefore the residuals should have a mean of zero, constant variance and ideally they are normally distributed. 



<!-- --------------------------------------------------------------------------------------------- -->
\pagebreak 

## Model 2: Seasonal Naive Method
Each forecast is equal to the last value from the same season.
In this case here the forecast values for 2013 are just the same values as from the year 2012.
Same day, same month, just the year is replaced.

The prediction provided by this model for the year 2013 is shown as blue line: 

```{r m2PlotForecast, fig.align='center', echo=FALSE, fig.height=2.8, out.width="100%", fig.show="hold"}
tsJenaTd.fc_snaive <- snaive(tsJenaTd.train, h=365)
autoplot(tsJenaTd.fc_snaive) + myggtheme +
  xlab("Date") +
  ylab("T (degC)") +
  ggtitle("Model 2: Seasonal naive method, forecast of \"T (degC)\"") +
  coord_cartesian(ylim = c(-25,40))
```

With this model the following accuracies on training and test dataset can be achieved: 

```{r m2TableAccuracy, fig.align='center', echo=FALSE, out.width="100%", fig.show="hold"}
# show accuracy of forecast model:
accuracy(tsJenaTd.fc_snaive, tsJenaTd.test) %>%
  kable() %>% kable_styling(bootstrap_options = "striped", latex_options = "striped", full_width = FALSE, position = "center")

# gather accuracies of forecast models:
accuracyTb.train <- bind_rows(accuracyTb.train, add_column(as_tibble(t(accuracy(tsJenaTd.fc_snaive, tsJenaTd.test)[1,])), Model="M2 seasonal naive", .before = 1))
accuracyTb.test  <- bind_rows(accuracyTb.test , add_column(as_tibble(t(accuracy(tsJenaTd.fc_snaive, tsJenaTd.test)[2,])), Model="M2 seasonal naive", .before = 1))
```

The residual distribution of this model is shown by a histogram:

```{r m2PlotResiduals, fig.align='center', echo=FALSE, fig.height=2.8, out.width="100%", fig.show="hold"}
# Plot residuals
res.df <- na.omit(data.frame(res = as.matrix(tsJenaTd.fc_snaive$residuals)))
myhistogram(res.df, res.df$res, 1, "Residuals of model 2: seasonal naive method" , "Residuals") + coord_cartesian(xlim = c(-25,20))
```



<!-- --------------------------------------------------------------------------------------------- -->
\pagebreak 

## Model 3: Average Method
```{r m3Forecast, echo=FALSE}
tsJenaTd.fc_meanf <- meanf(tsJenaTd.train, h=365)
```
The average model calculates the mean of the given historical data.
All forecasts are constant and equal to this mean value.
Here in this case it turns out, that the calculated average value is `r tsJenaTd.fc_meanf$mean[1]`, 
which is very similar to the last observation used for the naive model (`r tsJenaTd.fc_naive$mean[1]`).

The prediction provided by this model for the year 2013 is shown as blue line: 

```{r m3PlotForecast, fig.align='center', echo=FALSE, fig.height=2.8, out.width="100%", fig.show="hold"}
autoplot(tsJenaTd.fc_meanf) + myggtheme +
  xlab("Date") +
  ylab("T (degC)") +
  ggtitle("Model 3: Average method, forecast of \"T (degC)\"") +
  coord_cartesian(ylim = c(-25,40))
```

With this model the following accuracies on training and test dataset can be achieved: 

```{r m3TableAccuracy, fig.align='center', echo=FALSE, out.width="100%", fig.show="hold"}
# show accuracy of forecast model:
accuracy(tsJenaTd.fc_meanf , tsJenaTd.test) %>%
  kable() %>% kable_styling(bootstrap_options = "striped", latex_options = "striped", full_width = FALSE, position = "center")

# gather accuracies of forecast models:
accuracyTb.train <- bind_rows(accuracyTb.train, add_column(as_tibble(t(accuracy(tsJenaTd.fc_meanf , tsJenaTd.test)[1,])), Model="M3 average"       , .before = 1))
accuracyTb.test  <- bind_rows(accuracyTb.test , add_column(as_tibble(t(accuracy(tsJenaTd.fc_meanf , tsJenaTd.test)[2,])), Model="M3 average"       , .before = 1))
```

The residual distribution of this model is shown by a histogram:

```{r m3PlotResiduals, fig.align='center', echo=FALSE, fig.height=2.8, out.width="100%", fig.show="hold"}
# Plot residuals
res.df <- na.omit(data.frame(res = as.matrix(tsJenaTd.fc_meanf$residuals )))
myhistogram(res.df, res.df$res, 1, "Residuals of model 3: average method", "Residuals") + coord_cartesian(xlim = c(-25,20))
```



<!-- --------------------------------------------------------------------------------------------- -->
\pagebreak 

## Model 4: Drift Method
The drift method is best described as a forecast equal to last value plus average change.
It calculates the difference between first and last observation and extrapolates that gradient into the future.
Here the slope is small as there is no obvious trend present.

The prediction provided by this model for the year 2013 is shown as blue line: 

```{r m4PlotForecast, fig.align='center', echo=FALSE, fig.height=2.8, out.width="100%", fig.show="hold"}
tsJenaTd.fc_drift <- rwf(tsJenaTd.train, h=365, drift=T)
autoplot(tsJenaTd.fc_drift) + myggtheme +
  xlab("Date") +
  ylab("T (degC)") +
  ggtitle("Model 4: Drift method, forecast of \"T (degC)\"") +
  coord_cartesian(ylim = c(-25,40))
```

With this model the following accuracies on training and test dataset can be achieved: 

```{r m4TableAccuracy, fig.align='center', echo=FALSE, out.width="100%", fig.show="hold"}
# show accuracy of forecast model:
accuracy(tsJenaTd.fc_drift , tsJenaTd.test) %>%
  kable() %>% kable_styling(bootstrap_options = "striped", latex_options = "striped", full_width = FALSE, position = "center")

# gather accuracies of forecast models:
accuracyTb.train <- bind_rows(accuracyTb.train, add_column(as_tibble(t(accuracy(tsJenaTd.fc_drift , tsJenaTd.test)[1,])), Model="M4 drift"         , .before = 1))
accuracyTb.test  <- bind_rows(accuracyTb.test , add_column(as_tibble(t(accuracy(tsJenaTd.fc_drift , tsJenaTd.test)[2,])), Model="M4 drift"         , .before = 1))
```

The residual distribution of this model is shown by a histogram:

```{r m4PlotResiduals, fig.align='center', echo=FALSE, fig.height=2.8, out.width="100%", fig.show="hold"}
# Plot residuals
res.df <- na.omit(data.frame(res = as.matrix(tsJenaTd.fc_drift$residuals )))
myhistogram(res.df, res.df$res, 1, "Residuals of model 4: drift method"  , "Residuals") + coord_cartesian(xlim = c(-25,20))
```



<!-- --------------------------------------------------------------------------------------------- -->
\pagebreak 

## Model 5: Seasonal Decomposition + Naive Method
Seasonal decomposition means separating a seasonal time series into its constituent components, which are a seasonal component, a trend component and a random component.
The acronym *STL* is related to this method, i.e. decompose a time series into seasonal, trend and irregular components using loess.

Forecasts of STL objects are obtained by applying a non-seasonal forecasting method to the seasonally adjusted data and re-seasonalizing using the last year of the seasonal component.
Here the naive method is applied.

The prediction provided by this model for the year 2013 is shown as blue line: 

```{r m5PlotForecast, fig.align='center', echo=FALSE, fig.height=2.8, out.width="100%", fig.show="hold"}
tsJenaTd.fc_stl_naive   <- stlf(tsJenaTd.train, h=365, method = "naive")
autoplot(tsJenaTd.fc_stl_naive  ) + myggtheme +
  xlab("Date") +
  ylab("T (degC)") +
  ggtitle("Model 5: STL + naive method, forecast of \"T (degC)\"") +
  coord_cartesian(ylim = c(-25,40))
```

With this model the following accuracies on training and test dataset can be achieved: 

```{r m5TableAccuracy, fig.align='center', echo=FALSE, out.width="100%", fig.show="hold"}
# show accuracy of forecast model:
accuracy(tsJenaTd.fc_stl_naive  , tsJenaTd.test) %>%
  kable() %>% kable_styling(bootstrap_options = "striped", latex_options = "striped", full_width = FALSE, position = "center")

# gather accuracies of forecast models:
accuracyTb.train <- bind_rows(accuracyTb.train, add_column(as_tibble(t(accuracy(tsJenaTd.fc_stl_naive  , tsJenaTd.test)[1,])), Model="M5 STL + naive", .before = 1))
accuracyTb.test  <- bind_rows(accuracyTb.test , add_column(as_tibble(t(accuracy(tsJenaTd.fc_stl_naive  , tsJenaTd.test)[2,])), Model="M5 STL + naive", .before = 1))
```

The residual distribution of this model is shown by a histogram:

```{r m5PlotResiduals, fig.align='center', echo=FALSE, fig.height=2.8, out.width="100%", fig.show="hold"}
# Plot residuals
res.df <- na.omit(data.frame(res = as.matrix(tsJenaTd.fc_stl_naive$residuals  )))
myhistogram(res.df, res.df$res, 1, "Residuals of model 5: stl + naive method", "Residuals") + coord_cartesian(xlim = c(-25,20))
```



<!-- --------------------------------------------------------------------------------------------- -->
\pagebreak 

## Model 6: Seasonal Decomposition + Drift Method
This model uses the STL method like described before and applies the drift method to obtain a forecast.

The prediction provided by this model for the year 2013 is shown as blue line: 

```{r m6PlotForecast, fig.align='center', echo=FALSE, fig.height=2.8, out.width="100%", fig.show="hold"}
tsJenaTd.fc_stl_rwdrift <- stlf(tsJenaTd.train, h=365, method = "rwdrift")
autoplot(tsJenaTd.fc_stl_rwdrift) + myggtheme +
  xlab("Date") +
  ylab("T (degC)") +
  ggtitle("Model 6: STL + drift method, forecast of \"T (degC)\"") +
  coord_cartesian(ylim = c(-25,40))
```

With this model the following accuracies on training and test dataset can be achieved: 

```{r m6TableAccuracy, fig.align='center', echo=FALSE, out.width="100%", fig.show="hold"}
# show accuracy of forecast model:
accuracy(tsJenaTd.fc_stl_rwdrift, tsJenaTd.test) %>%
  kable() %>% kable_styling(bootstrap_options = "striped", latex_options = "striped", full_width = FALSE, position = "center")

# gather accuracies of forecast models:
accuracyTb.train <- bind_rows(accuracyTb.train, add_column(as_tibble(t(accuracy(tsJenaTd.fc_stl_rwdrift, tsJenaTd.test)[1,])), Model="M6 STL + drift", .before = 1))
accuracyTb.test  <- bind_rows(accuracyTb.test , add_column(as_tibble(t(accuracy(tsJenaTd.fc_stl_rwdrift, tsJenaTd.test)[2,])), Model="M6 STL + drift", .before = 1))
```

The residual distribution of this model is shown by a histogram:

```{r m6PlotResiduals, fig.align='center', echo=FALSE, fig.height=2.8, out.width="100%", fig.show="hold"}
# Plot residuals
res.df <- na.omit(data.frame(res = as.matrix(tsJenaTd.fc_stl_rwdrift$residuals)))
myhistogram(res.df, res.df$res, 1, "Residuals of model 6: stl + drift method", "Residuals") + coord_cartesian(xlim = c(-25,20))
```

As there is no distinct drift present, the resulting values of this model look quite similar to
the values obtained by the *seasonal decomposition + naive model*.  



<!-- --------------------------------------------------------------------------------------------- -->
\pagebreak 

## Model 7: Seasonal Decomposition + ARIMA Method
This model uses the STL method and applies the ARIMA method to obtain a forecast.
The ARIMA model is described later as a separate method.

The prediction provided by this model for the year 2013 is shown as blue line: 

```{r m7PlotForecast, fig.align='center', echo=FALSE, fig.height=2.8, out.width="100%", fig.show="hold"}
tsJenaTd.fc_stl_arima   <- stlf(tsJenaTd.train, h=365, method = "arima")
autoplot(tsJenaTd.fc_stl_arima  ) + myggtheme +
  xlab("Date") +
  ylab("T (degC)") +
  ggtitle("Model 7: STL + ARIMA method, forecast of \"T (degC)\"") +
  coord_cartesian(ylim = c(-25,40))
```

With this model the following accuracies on training and test dataset can be achieved: 

```{r m7TableAccuracy, fig.align='center', echo=FALSE, out.width="100%", fig.show="hold"}
# show accuracy of forecast model:
accuracy(tsJenaTd.fc_stl_arima  , tsJenaTd.test) %>%
  kable() %>% kable_styling(bootstrap_options = "striped", latex_options = "striped", full_width = FALSE, position = "center")

# gather accuracies of forecast models:
accuracyTb.train <- bind_rows(accuracyTb.train, add_column(as_tibble(t(accuracy(tsJenaTd.fc_stl_arima  , tsJenaTd.test)[1,])), Model="M7 STL + ARIMA", .before = 1))
accuracyTb.test  <- bind_rows(accuracyTb.test , add_column(as_tibble(t(accuracy(tsJenaTd.fc_stl_arima  , tsJenaTd.test)[2,])), Model="M7 STL + ARIMA", .before = 1))
```

The residual distribution of this model is shown by a histogram:

```{r m7PlotResiduals, fig.align='center', echo=FALSE, fig.height=2.8, out.width="100%", fig.show="hold"}
# Plot residuals
res.df <- na.omit(data.frame(res = as.matrix(tsJenaTd.fc_stl_arima$residuals  )))
myhistogram(res.df, res.df$res, 1, "Residuals of model 7: stl + ARIMA method", "Residuals") + coord_cartesian(xlim = c(-25,20))
```

The residuals are very close to a normal distribution.



<!-- --------------------------------------------------------------------------------------------- -->
\pagebreak 

## Model 8: Seasonal ARIMA Method
The [Autoregressive Integrated Moving Average (ARIMA) model](https://en.wikipedia.org/wiki/Autoregressive_integrated_moving_average)
for univariate time series combines three parts
with the integers ($p$,$d$,$q$) denoting the grade of order of the three parts:

* **AR** - autoregressive part ($p$)

* **I** - integration, i.e. degree of differencing ($d$)

* **MA** - moving average part ($q$)

For non-stationary time series differencing is applied $d$-times to obtain a stationary signal.
This stationary signal is then represented by the *ARMA*-model.
To obtain the values for the original non-stationary signal again, integration is performed on the results of the *ARMA*-model.
The *ARMA*-model is defined as:
$$y_t = c + \epsilon_t + \sum_{i=1}^p a_i y_{t-i} + \sum_{j=1}^q b_j \epsilon_{t-j}$$
The signal $y_t$ is build by a weighted moving average of random values $\epsilon_{t-j}$, $j=1,...,q$ of the $q$ previous periods. 
The so-called **MA**-coefficients $b_j, j=1,...,q$ determine with which weight the random values contribute to the signal.
Furthermore the signal is composed of a constant $c$, a random value $\epsilon_t$ and
a weighted moving average of the previous $p$ signal values $y_{t-i}$,
where the **AR**-coefficients $a_i, i=1,...,p$ are the weights.
Regarding the random values $\epsilon_t$,
it is assumed that they are time independent of each other and are normally distributed.

The goal is to determine the three describing parameters ($p$,$d$,$q$) of the *ARIMA*-model such that
the *AIC*-value is as small as possible.
The [*Akaike Information Criterion (AIC)*](https://en.wikipedia.org/wiki/Akaike_information_criterion)
is a widely used measure to quantify the fitting quality together with the simplicity of the model.
The R-function *[auto.arima()](https://www.rdocumentation.org/packages/forecast/versions/8.7/topics/auto.arima)*
uses the *AIC*-value to automatically determine the "best" parameters ($p$,$d$,$q$) for a given dataset.

For the training dataset the 
[autocorrelation function](https://en.wikipedia.org/wiki/Autocorrelation) 
(*ACF*) and the 
[partial autocorrelation function](https://en.wikipedia.org/wiki/Partial_autocorrelation_function)
(*PACF*) show several bars ranging out of the 95% confidence intervals (blue lines).

```{r m8ggtsdisplay, fig.align='center', echo=FALSE, fig.height=3.5, out.width="100%", fig.show="hold"}
ggtsdisplay(tsJenaTd.train, lag.max = 36, theme = theme_bw())
```

There is a good chance that an *ARIMA*-model is able to reconstruct and forecast the signal with high accuracy, as...

* *ACF* is used to identify the moving average (MA) part of the ARIMA model (omit first bar)

* *PACF* is used to identify the autoregressive part (AR)


\pagebreak 

The prediction provided by the ARIMA model for the year 2013 is shown as blue line: 

```{r m8PlotForecast, fig.align='center', echo=FALSE, fig.height=2.8, out.width="100%", fig.show="hold", results = 'hide', message=FALSE, warning=FALSE}
# model 8: seasonal ARIMA. explaination e.g.: https://otexts.com/fpp2/seasonal-arima.html
tsJenaTd.mo_auto.arima1 <- auto.arima(tsJenaTd.train, trace = T, seasonal = T) # ARIMA(5,0,0)(0,1,0)[365]
# better results with: approximation = F & stepwise = F , but very time-consuming!!!

tsJenaTd.fc_auto.arima1 <- forecast(tsJenaTd.mo_auto.arima1, h=365)

autoplot(tsJenaTd.fc_auto.arima1) + myggtheme +
  xlab("Date") +
  ylab("T (degC)") +
  ggtitle("Model 8: seasonal ARIMA method, forecast of \"T (degC)\"") +
  coord_cartesian(ylim = c(-25,40))
```

With this model the following accuracies on training and test dataset can be achieved: 

```{r m8TableAccuracy, fig.align='center', echo=FALSE, out.width="100%", fig.show="hold"}
# show accuracies of forecast model:
accuracy(tsJenaTd.fc_auto.arima1, tsJenaTd.test) %>%
  kable() %>% kable_styling(bootstrap_options = "striped", latex_options = "striped", full_width = FALSE, position = "center")

# gather accuracies of forecast models:
accuracyTb.train <- bind_rows(accuracyTb.train, add_column(as_tibble(t(accuracy(tsJenaTd.fc_auto.arima1, tsJenaTd.test)[1,])), Model="M8 seasonal ARIMA", .before = 1))
accuracyTb.test  <- bind_rows(accuracyTb.test , add_column(as_tibble(t(accuracy(tsJenaTd.fc_auto.arima1, tsJenaTd.test)[2,])), Model="M8 seasonal ARIMA", .before = 1))
```

The residual distribution of this model is shown by a histogram:

```{r m8PlotResiduals, fig.align='center', echo=FALSE, fig.height=2.8, out.width="100%", fig.show="hold"}
# Plot residuals
res.df <- na.omit(data.frame(res = as.matrix(tsJenaTd.fc_auto.arima1$residuals  )))
myhistogram(res.df, res.df$res, 1, "Residuals of model 8: seasonal ARIMA method", "Residuals") + coord_cartesian(xlim = c(-25,20))
```

The residual distribution shows a huge peak for the zero bin. 
This might be an indication for overfitting the given data.
At least the goal of obtaining a normal distribution is not achieved here. 



<!-- --------------------------------------------------------------------------------------------- -->
\pagebreak 

## Model 9: Neural Net Method 
A [neural network](https://en.wikipedia.org/wiki/Artificial_neural_network)
simplifies input variables over one or several layers into one output value.
In this case a *Neural Network Autoregression Model* is used with $p$ lagged values as input and
$k$ nodes in the hidden layer.
The R function [nnetar](https://www.rdocumentation.org/packages/forecast/versions/8.7/topics/nnetar)
determines automatically the parameters $p$ and $k$.


The prediction provided by this model for the year 2013 is shown as blue line: 

```{r m9PlotForecast, fig.align='center', echo=FALSE, fig.height=2.8, out.width="100%", fig.show="hold", results = 'hide', message=FALSE, warning=FALSE}
# model 9: Neural Net method, NNAR(p,P,k)  P=1 1-seasonal-lag (9,1,5)
tsJenaTd.mo_nnetar <- nnetar(tsJenaTd.train)

tsJenaTd.mo_nnetar # NNAR(5,1,4)[365] : 6-4-1 network with 33 weights - linear output units 

tsJenaTd.fc_nnetar <- forecast(tsJenaTd.mo_nnetar, h=365, PI=T)

autoplot(tsJenaTd.fc_nnetar) + myggtheme +
  xlab("Date") +
  ylab("T (degC)") +
  ggtitle("Model 9: Neural Net method, forecast of \"T (degC)\"") +
  coord_cartesian(ylim = c(-25,40))
```

With this model the following accuracies on training and test dataset can be achieved: 

```{r m9TableAccuracy, fig.align='center', echo=FALSE, out.width="100%", fig.show="hold"}
# show accuracy of forecast model:
accuracy(tsJenaTd.fc_nnetar, tsJenaTd.test) %>%
  kable() %>% kable_styling(bootstrap_options = "striped", latex_options = "striped", full_width = FALSE, position = "center")

# gather accuracies of forecast models:
accuracyTb.train <- bind_rows(accuracyTb.train, add_column(as_tibble(t(accuracy(tsJenaTd.fc_nnetar, tsJenaTd.test)[1,])), Model="M9 Neural Net", .before = 1))
accuracyTb.test  <- bind_rows(accuracyTb.test , add_column(as_tibble(t(accuracy(tsJenaTd.fc_nnetar, tsJenaTd.test)[2,])), Model="M9 Neural Net", .before = 1))
```

The residual distribution of this model is shown by a histogram:

```{r m9PlotResiduals, fig.align='center', echo=FALSE, fig.height=2.8, out.width="100%", fig.show="hold"}
# Plot residuals
res.df <- na.omit(data.frame(res = as.matrix(tsJenaTd.fc_nnetar$residuals  )))
myhistogram(res.df, res.df$res, 1, "Residuals of model 9: Neural Net method", "Residuals") + coord_cartesian(xlim = c(-25,20))
```



<!-- --------------------------------------------------------------------------------------------- -->
\pagebreak 

# Results

Following table summarises the achieved accuracy values for fitting the training data:

```{r tableAccuracyTraining, fig.align='center', echo=FALSE, out.width="100%", fig.show="hold"}
# define scale function for better visualisation
scale01 <- function(x){((x-min(x))/(max(x)-min(x))*0.88*10 + 0.12*10)*1}

# take a look on the accuracy overview (train dataset):
accuracyTb.train %>% select(Model,RMSE,MAE,MAPE,MASE) %>% kable() %>% kable_styling(bootstrap_options = "striped", latex_options = "striped", full_width = FALSE, position = "center")
```

Visualization of accuracy as *Scaled Error* value for training dataset:

```{r plotAccuracyTraining, fig.align='center', echo=FALSE, fig.height=3.5, out.width="100%", fig.show="hold"}
# scale the train dataset errors and unpivot the table for plotting purpose:
df.accuracyTb01.train <- cbind(select(accuracyTb.train, Model),
                        lapply(select(accuracyTb.train, RMSE,MAE,MAPE,MASE), scale01) ) %>% 
  melt(id = c("Model")) %>% rename(Error = variable)

# plot Errors overview for train dataset:
ggplot(data=df.accuracyTb01.train, aes(x=reorder(Model, desc(Model)), y=value, fill=Error)) +
  geom_bar(stat="identity", position=position_dodge()) +
  scale_y_continuous(trans = 'log10', breaks = c((1:10)), labels = c("small","","","","","","","","","big")) +
  xlab("") +
  ylab("Scaled Error Value") +
  coord_flip() +
  ggtitle("Scaled Error value for training dataset: Which model achieves the lowest Error") +
  myggtheme +
  theme(panel.grid.minor = element_blank())
```

Model *M7*, the *Seasonal Decomposition + ARIMA Method* shows the best accuracy for fitting the training data.


\pagebreak 

Following table summarises the achieved accuracy values for forecasting the test data:

```{r tableAccuracyTest, fig.align='center', echo=FALSE, out.width="100%", fig.show="hold"}
# take a look on the accuracy overview (test dataset):
accuracyTb.test  %>% select(Model,RMSE,MAE,MAPE,MASE) %>% kable() %>% kable_styling(bootstrap_options = "striped", latex_options = "striped", full_width = FALSE, position = "center")
```

Visualization of accuracy as *Scaled Error* value for test dataset:

```{r plotAccuracyTest, fig.align='center', echo=FALSE, fig.height=3.5, out.width="100%", fig.show="hold"}
# scale the test dataset errors and unpivot the table for plotting purpose:
df.accuracyTb01.test <- cbind(select(accuracyTb.test, Model),
                       lapply(select(accuracyTb.test, RMSE,MAE,MAPE,MASE), scale01) ) %>% 
  melt(id = c("Model")) %>% rename(Error = variable)

# plot Errors overview for test dataset:
ggplot(data=df.accuracyTb01.test, aes(x=reorder(Model, desc(Model)), y=value, fill=Error)) +
  #ggplot(data=df.accuracyTb01.test, aes(x=Error, y=value, fill=Model)) +
  geom_bar(stat="identity", position=position_dodge()) +
  #  scale_y_log10(limits = c(0.1,100)) +
  scale_y_continuous(trans = 'log10', breaks = c((1:10)), labels = c("small","","","","","","","","","big")) +
  xlab("") +
  ylab("Scaled Error Value") +
  coord_flip() +
  ggtitle("Scaled Error value for test dataset: Which model achieves the lowest Error") +
  myggtheme +
  theme(panel.grid.minor = element_blank())
```

Model *M9*, the *Neural Net Method* provides the best accuracy regarding forecasting the test data.
Also the seasonal methods *M2 - Seasonal Naive Method* and *M8 - Seasonal ARIMA Method*
do a quite good job.
The three models were not the favorites when looking on the accuracy for fitting the training dataset. 

The following plots visualize the fitted and forecasted values of the models M7, M9 and M8: 

```{r plotForecastTestM7, fig.align='center', echo=FALSE, fig.height=3.1, out.width="100%", fig.show="hold", message=FALSE, warning=FALSE}
# plot forecast vs. test data for M7
ggplot() +
  forecast::autolayer(tsJenaTd.train, series = 'train data', size = 1) +
  forecast::autolayer(tsJenaTd.test , series = 'test data' , size = 1) +
  forecast::autolayer(tsJenaTd.fc_stl_arima$fitted, series = 'M7 fitted', size = 0.5) +
  forecast::autolayer(tsJenaTd.fc_stl_arima$mean  , series = 'M7 forecast', size = 0.5) +
  xlab('Year') + ylab('daily mean of \"T (degC)\"') + 
  guides(colour = guide_legend(title = 'Dataset', reverse=F)) +
  myggtheme +
  coord_cartesian(ylim = c(-25,40)) +
  #theme(legend.position = c(0.85, 0.13)) +
  #theme(legend.direction = "horizontal", legend.position = c(0.35, 0.067)) +
  theme(legend.direction = "horizontal", legend.position = c(0.35, 0.9)) +
  scale_color_manual(values=c("cyan", "blue", "green","black"),
                     breaks=c("train data", "test data", "M7 fitted", "M7 forecast"))
```


```{r plotForecastTestM9, fig.align='center', echo=FALSE, fig.height=3.1, out.width="100%", fig.show="hold", message=FALSE, warning=FALSE}
# plot forecast vs. test data for M9
ggplot() +
  forecast::autolayer(tsJenaTd.train, series = 'train data', size = 1) +
  forecast::autolayer(tsJenaTd.test , series = 'test data' , size = 1) +
  forecast::autolayer(tsJenaTd.fc_nnetar$fitted, series = 'M9 fitted'  , size = 0.5) +
  forecast::autolayer(tsJenaTd.fc_nnetar$mean  , series = 'M9 forecast', size = 0.5) +
  xlab('Year') + ylab('daily mean of \"T (degC)\"') + 
  guides(colour = guide_legend(title = 'Dataset', reverse=F)) +
  myggtheme +
  coord_cartesian(ylim = c(-25,40)) +
  #theme(legend.position = c(0.85, 0.13)) +
  theme(legend.direction = "horizontal", legend.position = c(0.35, 0.9)) +
  scale_color_manual(values=c("cyan", "blue", "green","black"),
                     breaks=c("train data", "test data", "M9 fitted", "M9 forecast"))
```


```{r plotForecastTestM8, fig.align='center', echo=FALSE, fig.height=3.1, out.width="100%", fig.show="hold", message=FALSE, warning=FALSE}
# plot forecast vs. test data for M8
ggplot() +
  forecast::autolayer(tsJenaTd.train, series = 'train data', size = 1) +
  forecast::autolayer(tsJenaTd.test , series = 'test data' , size = 1) +
  forecast::autolayer(tsJenaTd.fc_auto.arima1$fitted, series = 'M8 fitted'  , size = 0.5) +
  forecast::autolayer(tsJenaTd.fc_auto.arima1$mean  , series = 'M8 forecast', size = 0.5) +
  xlab('Year') + ylab('daily mean of \"T (degC)\"') + 
  guides(colour = guide_legend(title = 'Dataset', reverse=F)) +
  myggtheme +
  coord_cartesian(ylim = c(-25,40)) +
  #theme(legend.position = c(0.85, 0.13)) +
  theme(legend.direction = "horizontal", legend.position = c(0.35, 0.9)) +
  scale_color_manual(values=c("cyan", "blue", "green","black"),
                     breaks=c("train data", "test data", "M8 fitted", "M8 forecast"))
```



<!-- --------------------------------------------------------------------------------------------- -->

\pagebreak 

# Conclusion

In this project machine learning algorithms were used to forecast the daily mean temperatures of one year by fitting the models to the available data of the previous four years.
The quality of this task was quantified by a selection of accuracy measurements like RMSE, MAE, MASE and MAPE.

A variety of models were used:
Some of the simple ones like naive-, average- or drift-methods already achieve good accuracy.
This might be because a large part of the signal (=temperatures) is noise, which usually cannot be forecasted well by sophisticated models.
It is also obvious that the signal shows high seasonality.
Therefore models based on seasonal decomposition (STL) show quite good performance.
Another standard tool is the ARIMA-model.
In the simple form, like it was used here, it tended to *overfitt* the training data.
So the forecast accurcacy was dissapointing. 
The neural net model on the other hand fitted the training data with fair quality and showed best forecast accuracy. 

It is likely that much more sophisticated methods would achieve better accuracy.
So far this project was just a first survey of a selection of methods without any detailed investigation or optimization of each used model. 
Furthermore the validation by dividing the data into one training and one test dataset could be improved.
For example an approach like cross-validation could help to avoid overfitting.
For simplicity reasons it was not applied so far.
The implementation is left to future work.


<!-- --------------------------------------------------------------------------------------------- -->
<!-- \pagebreak --> 

# Appendix - Sessioninfo
```{r , echo=FALSE, size = 'tiny'}
sessionInfo()
```
